---
title: "Data prep"
output: html_notebook
---

# Purpose

Prepare results from the KCHD 2024 Mental Health Survey for analysis.

# Set up environment and import data

```{r message=FALSE}
library(tidyverse)
library(readxl)
library(sf)
```

```{r}
source("fn.R")
```

Import raw survey results from ETC.

```{r}
mhsurv <- read_excel("../data/1-source/KC Health_raw data_Mental Health Survey Results 2024.xls")
```

Import codebook.

```{r}
sh <- excel_sheets("../data/1-source/mh-codebook.xlsx")
cb <- lapply(sh, \(x) read_excel(
  "../data/1-source/mh-codebook.xlsx",
  sheet = x
))
names(cb) <- sh
```

Import HOI regions.

```{r}
hoi_regions <- st_read("../data/1-source/hoi_regions.shp", quiet = TRUE)
```

# Prep data

Rename variables.

```{r}
colnames(mhsurv) <- cb$key$nm2[cb$key$in_src]
```

## Validate

### `id`

Validate `id`.

```{r}
any(duplicated(mhsurv$id))
```

Check if `id` & `id2` are identical.

```{r}
identical(mhsurv$id, mhsurv$id2)
```

### Demographics

Missing or invalid race/ethnicity values.

```{r}
mhsurv |>
  filter(if_all(race_ethn_01:race_ethn_08, ~ !.x %in% cb$race_ethn$number)) |>
  nrow()
```

Missing or invalid birth sex values.

```{r}
mhsurv |>
  filter(!birth_sex %in% cb$birth_sex$number) |>
  nrow()
```

Missing or invalid current gender values.

```{r}
mhsurv |>
  filter(if_all(cur_gen_01:cur_gen_06, ~ !.x %in% cb$cur_gen$number)) |>
  nrow()
```

Missing or invalid age values.

```{r}
mhsurv |>
  filter(is.na(age1) | !age1 %in% cb$age1$number) |>
  nrow()
```

### Questions

Validate question data against expected values in the codebook.

```{r}
key <- cb$key

vars <- key$nm2[grepl("^race_ethn_\\d|^birth_|^cur_gen_\\d|age1|^q\\d", key$nm2)]

ls <- lapply(vars, \(x) {
  sh <- key[which(key$nm2 == x), "code_val", drop = TRUE]
  vals <- unique(mhsurv[[x]])
  vals[!vals %in% cb[[sh]]$number]
})

names(ls) <- vars
```

Q15: number of hours per day.

```{r}
mhsurv |>
  filter(q15 < 0 | q15 > 24) |>
  nrow()
```

Q17: number of minutes to fall asleep.

```{r}
mhsurv |>
  filter(q17 < 0 | q17 > 1200) |>
  nrow()
```

Q19: number of hours slept per day.

```{r}
mhsurv |>
  filter(q19 < 0 | q19 > 24) |>
  nrow()
```

Reversed scale question.

```{r}
mhsurv |>
  select(id, starts_with("q13_")) |>
  bind_cols(n_unq = apply(
    mhsurv |>
      select(starts_with("q13_")),
    1,
    function(x) length(unique(x))
  )) |>
  filter(if_all(starts_with("q13_"), ~ .x %in% 1:5)) |>
  filter(n_unq == 1)
```

Questions 16 and 18 (time of day) validated in scoring section below.

## Codebook

Convert question data to factor.

```{r}
# Factorize as-is
sh <- unique(key[which(key$factor_dir == "asis"), "code_score", drop = TRUE])
cb[sh] <- lapply(sh, \(x) {
  cb[[x]] |>
    mutate(text = factor(text, levels = text))
})

# Factorize in reverse
sh <- unique(key[which(key$factor_dir == "rev"), "code_score", drop = TRUE])
cb[sh] <- lapply(sh, \(x) {
  cb[[x]] |>
    mutate(text = factor(text, levels = rev(text)))
})
```

# Reconfigure sample characteristics

## Race/ethnicity

Combine multiple answers (`race_ethn_combined`).

```{r}
# Create list column for race/ethnicity values
mhsurv <- mhsurv |>
  rowwise() |>
  mutate(
    race_ethn_n = list(c_across(matches("race_ethn_\\d"))),
    .after = race_ethn_other
  ) |>
  ungroup()

# Remove NAs from lists
mhsurv$race_ethn_n <- lapply(mhsurv$race_ethn_n, \(x) x[!is.na(x)])

# Replace numbers with text from codebook
mhsurv$race_ethn_t1 <- lapply(mhsurv$race_ethn_n, \(x) {
  r <- which(cb$race_ethn1$number %in% x)
  cb$race_ethn1$text[r]
})
mhsurv$race_ethn_t2 <- lapply(mhsurv$race_ethn_n, \(x) {
  r <- which(cb$race_ethn2$number %in% x)
  cb$race_ethn2$text[r]
})

# Combine lists into strings
mhsurv$race_ethn_combined1 <- lapply(mhsurv$race_ethn_t1, \(x) {
 paste(x, collapse = "; ")
})
mhsurv$race_ethn_combined2 <- lapply(mhsurv$race_ethn_t2, \(x) {
 paste(x, collapse = "; ")
})

# Replace blanks with NA
mhsurv <- mhsurv |>
  rowwise() |>
  mutate(
    race_ethn_combined1 = unlist(if_else(
      race_ethn_combined1 == "", NA, race_ethn_combined1
    )),
    race_ethn_combined2 = unlist(if_else(
      race_ethn_combined2 == "", NA, race_ethn_combined2
    ))
  ) |>
  ungroup() |>
  relocate(
    race_ethn_t1, race_ethn_combined1,
    race_ethn_t2, race_ethn_combined2,
    .after = race_ethn_n
  )
```

Create collapsed category (`race_ethn_bwo`).

- "Black" includes anyone who selected only "African-American/Black" or "African".
- "White" includes anyone who selected only "Middle Eastern or Northern African" or "White/Caucasian".
- "Other" includes anyone who selected another category alone and anyone who selected multiple categories.

```{r}
mhsurv <- mhsurv |>
  mutate(
    race_ethn_bwo = case_when(
      grepl("Asian|Hispanic|IndAm|Other|;", race_ethn_combined2) ~ "Other",
      grepl("AfAm|African", race_ethn_combined2) ~ "Black",
      grepl("MENA|White", race_ethn_combined2) ~ "White"
    ),
    .after = race_ethn_combined2
  ) |>
  mutate(race_ethn_bwo = factor(
    race_ethn_bwo,
    levels = c("Black", "White", "Other")
  ))
```

## Hispanic origin

Hispanic origin is a variable created from responses to the race/ethnicity question to align with the Hispanic origin category used by the US Census that is distinct from race.

- "Hispanic" includes anyone who selected "Hispanic/Latinx" alone or in combination with another category.
- "Not Hispanic" includes anyone who did not select "Hispanic/Latinx".

```{r}
mhsurv <- mhsurv |>
  mutate(
    hispanic_origin = if_else(
      race_ethn_04 == 4, "Hispanic", "Not Hispanic", missing = "Not Hispanic"
    ),
    .after = race_ethn_bwo
  ) |>
  mutate(hispanic_origin = factor(
    hispanic_origin,
    levels = c("Hispanic", "Not Hispanic")
  ))
```

## Current gender

Combine multiple answers (`cur_gen_combined`).

```{r}
# Create list column for race/ethnicity values
mhsurv <- mhsurv |>
  rowwise() |>
  mutate(
    cur_gen_n = list(c_across(matches("cur_gen_\\d"))),
    .after = cur_gen_self_id
  ) |>
  ungroup()

# Remove NAs from lists
mhsurv$cur_gen_n <- lapply(mhsurv$cur_gen_n, \(x) x[!is.na(x)])

# Replace numbers with text from codebook
mhsurv$cur_gen_t <- lapply(mhsurv$cur_gen_n, \(x) {
  r <- which(cb$cur_gen$number %in% x)
  cb$cur_gen$text[r]
})

# Combine lists into strings
mhsurv$cur_gen_combined <- lapply(mhsurv$cur_gen_t, \(x) {
 paste(x, collapse = "; ")
})

# Replace blanks with NA
mhsurv <- mhsurv |>
  rowwise() |>
  mutate(cur_gen_combined = unlist(if_else(
    cur_gen_combined == "", NA, cur_gen_combined
  ))) |>
  ungroup() |>
  relocate(cur_gen_t, cur_gen_combined, .after = cur_gen_n)
```

## Age

Duplicate `age1` to use both original (`age1`) and collapsed (`age2`) categories.

```{r}
mhsurv <- mhsurv |>
  mutate(
    age2 = if_else(age1 == 6, 5, age1),
    .after = age1
  )
```

## Lon/lat

Correct longitude and latitude variables, which lack a decimal in the raw data.

```{r}
mhsurv <- mhsurv |>
  mutate(across(c(block_lon, block_lat), ~ .x / 10^6))
```

## HOI region

Join HOI region variable.

```{r}
surv_rgn <- mhsurv |>
  select(id, block_lon, block_lat) |>
  st_as_sf(coords = c("block_lon", "block_lat"), crs = 4326) |>
  st_join(hoi_regions[, "region"])

mhsurv <- mhsurv |>
  bind_cols(
    surv_rgn |>
      as_tibble() |>
      select(region)
  ) |>
  mutate(region = factor(region, levels = c("North", "Central", "South"))) |>
  relocate(hoi_region = region, .after = block_lat)
```

# Recode & score questions

## Q8: Pre/post COVID-19

Group responses to Q8 as "Better", "About the same", or "Worse".

```{r}
mhsurv <- mhsurv |>
  mutate(
    q8_recode = case_when(
      q8 == 1 | q8 == 2 ~ 1, # Better
      q8 == 3           ~ 2, # About the same
      q8 == 4 | q8 == 5 ~ 3, # Worse
      .default = q8
    ),
    .after = q8
  )
```

## Q11: Needed but didn't get treatment

Q11 should only be completed if the answer to Q10 is `1` ("Yes").

Count responses to Q10.

```{r}
mhsurv |>
  mutate(number = q10) |>
  group_by(number) |>
  count() |>
  left_join(cb$yn, by = "number") |>
  relocate(text, .after = number)
```

Count rows where Q10 is `1` and any Q11 part is `NA`.

```{r}
mhsurv |>
  filter(q10 == 1) |>
  filter(if_any(q11_01:q11_10, ~ is.na(.x))) |>
  nrow()
```

Count rows where Q10 is not `1` and any Q11 part is not `NA`.

```{r}
mhsurv |>
  filter(q10 != 1) |>
  filter(if_any(q11_01:q11_10, ~ !is.na(.x))) |>
  nrow()
```

## Q13: Social isolation

### Explore

Count `NA`s.

```{r}
cols <- colnames(mhsurv)[grepl("^q13_\\d", colnames(mhsurv))]
apply(mhsurv[, cols], 2, \(x) sum(is.na(x)))
```

Count "Don't know"s (`9`s).

```{r}
apply(mhsurv[, cols], 2, \(x) sum(ifelse(x == 9, TRUE, FALSE)))
```

Count respondents without any "Don't know"s.

```{r}
mhsurv |>
  filter(!if_any(matches("^q13_\\d"), ~ .x == 9)) |>
  nrow()
```

Plot the number of "Don't know"s per respondent.

```{r}
ls$q13 <- data.frame(n_dontknow = apply(
  mhsurv[, cols], 1, \(x) sum(ifelse(x == 9, TRUE, FALSE))
))

ls$q13$n_answered <- 11 - ls$q13$n_dontknow
```

```{r fig.width=6}
ls$q13 |>
  ggplot(aes(n_dontknow)) +
  geom_histogram(binwidth = 1)
```

### Score

Reverse scale for questions 01, 04, 07, 08, and 11 so that `1` represents low social isolation and `5` represents high social isolation.

```{r}
mhsurv <- mhsurv |>
  mutate(across(
    c(q13_01, q13_04, q13_07, q13_08, q13_11),
    ~ case_when(
      .x == 1 ~ 5,
      .x == 2 ~ 4,
      .x == 4 ~ 2,
      .x == 5 ~ 1,
      .default = .x
    )
  ))
```

Sum the responses to each part to calculate a total social isolation score.

```{r}
mhsurv$q13_scored <- rowSums(mhsurv[, cols])
```

Drop scores in rows with any "Don't know"s.

```{r}
mhsurv <- mhsurv |>
  mutate(q13_scored = if_else(
    if_any(matches("^q13_\\d"), ~ .x == 9),
    NA,
    q13_scored
  ))
```

Assign range values.

```{r}
mhsurv <- mhsurv |>
  mutate(q13_binned = case_when(
    q13_scored >= 11 & q13_scored <= 25 ~ 1, # Low
    q13_scored >= 26 & q13_scored <= 40 ~ 2, # Medium
    q13_scored >= 41 & q13_scored <= 55 ~ 3, # High
    .default = NA
  )) |>
  relocate(q13_scored, q13_binned, .after = q13_11)
```

## Q14: Social media addiction

### Explore

Count `NA`s.

```{r}
cols <- colnames(mhsurv)[grepl("^q14_\\d", colnames(mhsurv))]
apply(mhsurv[, cols], 2, \(x) sum(is.na(x)))
```

Count "Don't know"s (`9`s).

```{r}
apply(mhsurv[, cols], 2, \(x) sum(ifelse(x == 9, TRUE, FALSE)))
```

Count respondents without any "Don't know"s.

```{r}
mhsurv |>
  filter(!if_any(matches("^q14_\\d"), ~ .x == 9)) |>
  nrow()
```

Plot the number of "Don't know"s per respondent.

```{r}
ls$q14 <- data.frame(n_dontknow = apply(
  mhsurv[, cols], 1, \(x) sum(ifelse(x == 9, TRUE, FALSE))
))

ls$q14$n_answered <- 6 - ls$q14$n_dontknow
```

```{r fig.width=6}
ls$q14 |>
  ggplot(aes(n_dontknow)) +
  geom_histogram(binwidth = 1)
```

### Score

Sum the responses to each part to calculate a total social media addiction score.

```{r}
mhsurv$q14_scored <- rowSums(mhsurv[, cols])
```

Drop scores in rows with any "Don't know"s.

```{r}
mhsurv <- mhsurv |>
  mutate(q14_scored = if_else(
    if_any(matches("^q14_\\d"), ~ .x == 9),
    NA,
    q14_scored
  ))
```

Assign range values.

```{r}
mhsurv <- mhsurv |>
  mutate(q14_binned = case_when(
    q14_scored >= 6 & q14_scored <= 13 ~ 1, # Low
    q14_scored >= 14 & q14_scored <= 22 ~ 2, # Medium
    q14_scored >= 23 & q14_scored <= 30 ~ 3, # High
    .default = NA
  )) |>
  relocate(q14_scored, q14_binned, .after = q14_6)
```

## Q15: Hours on social media

Assign range values.

```{r}
mhsurv <- mhsurv |>
  mutate(
    q15_scored = case_when(
      q15 == 0 ~ 1,
      q15 >= 1 & q15 <= 4 ~ 2,
      q15 >= 5 & q15 <= 8 ~ 3,
      q15 > 8 & q15 <= 24 ~ 4,
      .default = NA
    ),
    .after = q15
  )
```

## Q16/18/19: Habitual sleep efficiency

Convert time variables to HH:MM format.

```{r}
mhsurv <- mhsurv |>
  mutate(bed1 = number_to_time(q16), .after = q16) |>
  mutate(wake1 = number_to_time(q18), .after = q18)
```

Validate times.

```{r}
mhsurv |>
  mutate(
    val_bed = validate_time(bed1),
    val_wake = validate_time(wake1),
    .after = wake1
  ) |>
  filter(!val_bed | !val_wake) |>
  select(q16, bed1, q18, wake1, val_bed, val_wake)
```

Convert bed time and wake time pairs to date objects.

```{r}
ls <- times_to_dates(mhsurv$bed1, mhsurv$wake1)
names(ls) <- c("bed2", "wake2")

mhsurv <- mhsurv |>
  bind_cols(ls) |>
  relocate(bed2, .after = bed1) |>
  relocate(wake2, .after = wake1)
```

Validate times (relative to one another).

```{r}
mhsurv |>
  mutate(validate = as.numeric(difftime(wake2, bed2))) |>
  select(id, validate) |>
  filter(validate < 0)
```

Calculate the number of hours spent in bed.

```{r}
mhsurv$hrs_in_bed <- as.numeric(difftime(
  mhsurv$wake2,
  mhsurv$bed2,
  "hours"
))
```

Validate `hrs_in_bed`.

```{r}
any(mhsurv$hrs_in_bed < 0, na.rm = TRUE)
```

Calculate PSQI Component 4: Habitual sleep efficiency.

```{r}
mhsurv <- mhsurv |>
  mutate(hse_pct = q19 / hrs_in_bed * 100) |>
  mutate(hse_scored = score_hse(hse_pct)) |>
  relocate(hrs_in_bed, hse_pct, hse_scored, .after = wake2)
```

View percentages > 100.

```{r}
mhsurv |>
  filter(hse_pct > 100) |>
  select(q16:hse_scored)
```

Filter out percentages > 100.

```{r}
mhsurv <- mhsurv |>
  mutate(hse_scored = if_else(hse_pct > 100, NA, hse_scored))
```

## Q17: Time to fall asleep

Assign range values.

```{r}
mhsurv <- mhsurv |>
  mutate(q17 = case_when(
    q17 > 0 & q17 <= 10 ~ 1,
    q17 >= 11 & q17 <= 30 ~ 2,
    q17 >= 31 & q17 <= 60 ~ 3,
    q17 >= 61 & q17 <= 120 ~ 4,
    q17 >= 121 & q17 <= 3600 ~ 5,
    .default = NA
  ))
```

# Select data for analysis

```{r}
vars <- c(
  "id", "race_ethn_combined1", "race_ethn_combined2", "race_ethn_other",
  "race_ethn_bwo", "hispanic_origin", "birth_sex", "cur_gen_combined",
  "age1", "age2", "q4_1", "q4_2", "q4_3", "q4_4", "q5", "q6",
  "q7", "q8_recode", "q9", "q10", "q11_01", "q11_02", "q11_03", "q11_04",
  "q11_05", "q11_06", "q11_07", "q11_08", "q11_09", "q11_10", "q12",
  "q13_binned", "q14_binned", "q15_scored", "hse_scored", "q17", "zip",
  "district", "block_lon", "block_lat", "hoi_region"
)
```

```{r}
mhsurv2 <- mhsurv |>
  select(all_of(vars)) |>
  rename(
    q8 = q8_recode,
    q13_01 = q13_binned,
    q14_1 = q14_binned,
    q15 = q15_scored,
    q16 = hse_scored
  )

vars <- colnames(mhsurv2)
```

# Substitute text values

Recode question values from codebook.

```{r}
cols <- list()
cols[[1]] <- sort(colnames(mhsurv2))
```

```{r}
vars <- c(
  "birth_sex", "age1", "age2", "q4_1", "q4_2", "q4_3", "q4_4", "q5", "q6", "q7",
  "q8", "q9", "q10", "q11_01", "q11_02", "q11_03", "q11_04", "q11_05", "q11_06",
  "q11_07", "q11_08", "q11_09", "q11_10", "q12", "q13_01", "q14_1", "q15",
  "q16", "q17"
)
```

```{r}
ls <- lapply(vars, \(x) {
  sh <- key$code_score[which(key$nm2 == x)]
  alts <- setNames(cb[[sh]]$text, cb[[sh]]$number)
  unname(alts[mhsurv2[[x]]])
})
ls <- as.data.frame(ls, col.names = vars)
```

```{r}
mhsurv2 <- cbind(mhsurv2[, !colnames(mhsurv2) %in% vars], ls)
```

```{r}
# Reorder columns
vars <- list()
vars$demo <- c(
  "race_ethn_combined1", "race_ethn_combined2", "race_ethn_other",
  "race_ethn_bwo", "hispanic_origin", "birth_sex", "cur_gen_combined",
  "age1", "age2"
)
vars$geo <- c("zip", "district", "hoi_region", "block_lon", "block_lat")
vars$questions <- colnames(mhsurv2)[grepl("^q\\d", colnames(mhsurv2))]

mhsurv2 <- mhsurv2 |>
  select(id, all_of(vars$demo), all_of(vars$geo), all_of(vars$questions))

cols[[2]] <- sort(colnames(mhsurv2))
```

```{r}
if (!isTRUE(all.equal(cols[[1]], cols[[2]]))) {
  warning("Columns unequal after reordering")
}
```

# Save

```{r}
saveRDS(cb, "../data/2-final/mh_codebook.rds")
saveRDS(mhsurv2, "../data/2-final/mh_survey_results.rds")
```
