---
title: "Survey sample characteristics"
output: html_notebook
---

# Purpose

Summarize the demographic and geographic characteristics of the Mental Health Survey sample, and compare the demographic and geographic profile of the survey sample to census population estimates for Kansas City.

# Set up environment and import data

```{r setup}
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 3
)
```

```{r message=FALSE}
library(tidyverse)
library(readxl)
library(kcData)
library(setmeup)
library(sf)
```

```{r}
source("fn.R")
```

```{r}
# Survey results
cb <- readRDS("../data/2-final/mh_codebook.rds")
mhsurv <- readRDS("../data/2-final/mh_survey_results.rds")
```

```{r}
# HOI regions
hoi_regions <- st_read("../data/1-source/hoi_regions.shp", quiet = TRUE)
```

# Analysis

The survey was completed by 648 residents. The sample size is noted when it is below 648 for a particular question. Dark vertical lines on some bars indicate the estimated citywide percentage of the population category sampled. Citywide population estimates were taken from the 2023 5-Year American Community Survey conducted by the US Census Bureau.

```{r}
mhsamp <- list()
```

```{r}
# Line segment displacement
d <- .45
```

## Sex

### Summarize

Summarize the survey sample by sex.

```{r}
df <- mhsurv |>
  summarize_results(var = "birth_sex", na_rm = FALSE)
```

Pull the ACS population estimates by sex for Kansas City.

```{r}
kcpop <- list()

kcpop$total <- acs5_city_2023$estimate[acs5_city_2023$variable == "B01001_001"]

kcpop$sex <- acs5_city_2023 |>
  filter(variable %in% c("B01001_002", "B01001_026")) |>
  mutate(kc_total = kcpop$total)
```

Find the percentage of each sex category in KC sampled.

```{r}
df <- df |>
  left_join(
    kcpop$sex |>
      select(label, estimate, kc_total) |>
      mutate(label = str_remove_all(label, "Estimate|Total|[:punct:]")),
    by = c("birth_sex" = "label"),
    keep = TRUE # keeps `birth_sex` as factor
  ) |>
  select(-label) |>
  mutate(
    label = paste0(round_ties_away(pct, 0), "%"),
    pct_kc = pct(estimate, kc_total, digits = 5),
    d0 = as.numeric(birth_sex) - d,
    d1 = as.numeric(birth_sex) + d
  )

df
```

### Plot

Figure: Sex of survey sample. N = 647.

```{r}
figs <- list()

figs$sex1 <- df |>
  drop_na(birth_sex) |>
  ggplot(aes(x = pct, y = birth_sex, label = label)) |>
  mh_barplot(fill = "#F04C43", legend = FALSE) +
  scale_x_continuous(expand = c(0, NA)) +
  xlab("\nSample (%)") + ylab("Birth sex\n")

figs$sex1
```

```{r}
figs$sex2 <- figs$sex1 +
  kc_pop_segment(pct_kc, pct_kc, d0, d1)

figs$sex2
```

```{r}
figs$sex3 <- df |>
  drop_na(birth_sex) |>
  ggplot(aes(x = "", y = pct, label = label, fill = birth_sex)) |>
  mh_pie_chart(fill_colors = c("#318CCC", "#F04C43")) +
  theme(plot.margin = margin(-.25, -.2, -.25, -.45, "in"))

figs$sex3
```

```{r}
n <- sum(df$n[!is.na(df$birth_sex)])
figcap <- paste0(
  cb$prompts$caption[cb$prompts$number == 2],
  " N = ", n, "."
)

figcap
```

```{r}
mhsamp$sex <- list(
  sum = df,
  n = n,
  fig1 = figs$sex1,
  fig2 = figs$sex2,
  fig3 = figs$sex3,
  cap = figcap
)
```

## Age

### Summarize

Summarize the survey sample by age.

```{r}
df <- mhsurv |>
  summarize_results(var = "age1", na_rm = FALSE)
```

Pull the ACS population estimates by age for Kansas City.

```{r}
kcpop$age <- acs5_city_2023 |>
  filter(
    str_detect(label, paste0("(", paste(18:100, collapse = "|"), ").*years")),
    concept == "Sex by Age"
  ) |>
  mutate(kc_total = kcpop$total)
```

Summarize ACS estimates by the same age categories in the survey sample.

```{r}
kcpop$age <- kcpop$age |>
  select(label, estimate, kc_total) |>
  mutate(label = str_remove_all(label, "Estimate|Total|[:punct:]")) |>
  mutate(
    sex = str_extract(label, "^[:alpha:]+(?=\\d)"),
    age_group = str_extract(label, "\\d.*") |>
      str_remove("\\syears.*")
  )

rng <- list(18:34, 35:44, 45:54, 55:64, 65:74, 75:85)

kcpop$age <- data.frame(
  age = cb$age1$text,
  estimate = unlist(lapply(rng, \(x) {
    kcpop$age |>
      filter(str_detect(age_group, paste(x, collapse = "|"))) |>
      pull(estimate) |>
      sum()
  })),
  kc_total = unique(kcpop$age$kc_total)
)
```

Find the percentage of each age category in KC sampled.

```{r}
df <- df |>
  left_join(
    kcpop$age,
    by = c("age1" = "age")
  ) |>
  mutate(
    pct_kc = pct(estimate, kc_total, digits = 5),
    d0 = as.numeric(age1) - d,
    d1 = as.numeric(age1) + d
  )

df
```

### Plot

Figure: Age of survey sample. N = 609.

```{r}
figs$age1 <- df |>
  drop_na(age1) |>
  mutate(label = paste0(round_ties_away(pct, 0), "%")) |>
  ggplot(aes(x = pct, y = age1, label = label)) |>
  mh_barplot(fill = "#F04C43", legend = FALSE) +
  scale_x_continuous(expand = c(0, NA)) +
  xlab("\nSample (%)") + ylab("Age\n")

figs$age1
```

```{r}
figs$age2 <- figs$age1 +
  kc_pop_segment(pct_kc, pct_kc, d0, d1)

figs$age2
```

```{r}
n <- sum(df$n[!is.na(df$age1)])
figcap <- paste0(
  cb$prompts$caption[cb$prompts$number == 3],
  " N = ", n, "."
)

figcap
```

```{r}
mhsamp$age <- list(
  sum = df,
  n = n,
  fig1 = figs$age1,
  fig2 = figs$age2,
  cap = figcap
)
```

## Age structure

### Summarize

```{r}
df <- mhsurv |>
  count(birth_sex, age1) |>
  mutate(n = case_when(
    birth_sex == "Male" ~ -n,
    birth_sex == "Female" ~ n,
    .default = NA
  ))

df
```

### Plot

Figure: Age structure of survey sample. N = 608.

```{r fig.height=2.5, fig.width=6}
# Create symmetrical x-axis
n <- max(abs(df$n), na.rm = TRUE) + 5
rng <- c(-n, n)

figs$pyr1 <- df |>
  mutate(label = abs(n)) |>
  drop_na(birth_sex, age1, n) |>
  ggplot(aes(x = n, y = age1, label = label, fill = birth_sex)) |>
  mh_barplot(legend_title = TRUE) +
  xlab("\nSample (n)") + ylab("Age\n") +
  scale_x_continuous(labels = abs, limits = rng) +
  labs(fill = "Birth sex") +
  theme(panel.grid.major.y = element_line(color = "grey92")) +
  scale_fill_manual(values = c("#318CCC", "#F04C43"))

figs$pyr1
```

```{r}
n <- sum(abs(df$n[!is.na(df$age1)]), na.rm = TRUE)
figcap <- paste0(
  "Age structure of survey sample.",
  " N = ", n, "."
)

figcap
```

```{r}
mhsamp$age_sex <- list(
  sum = df,
  n = n,
  fig = figs$pyr1,
  cap = figcap
)
```

## Race/ethnicity (all)

### Summarize

Summarize the survey sample by all race/ethnicity options.

```{r}
# Split multiple selections into separate columns
df <- mhsurv |>
  select(race_ethn_combined2) |>
  separate_wider_delim(
    race_ethn_combined2,
    delim = "; ",
    names_sep = "_",
    too_few = "align_start"
  )

# Stack columns and summarize
df <- df |>
  stack(select = colnames(df)) |>
  select(race_ethn = values) |>
  drop_na() |>
  count(race_ethn, .drop = FALSE) |>
  mutate(pct = pct(n, 648, digits = 5))
```

```{r}
# Convert to factor sorted by size
lvl <- df |>
  arrange(pct) |>
  pull(race_ethn)

df <- df |>
  mutate(race_ethn = factor(race_ethn, levels = lvl))
```

Summarize the write-in responses for "Other".

```{r}
mhsurv |>
  count(race_ethn_other, .drop = FALSE)
```

### Plot

Figure: Race/ethnicity of survey sample. Multiple selections possible.

```{r fig.height=3.5}
figs$re1 <- df |>
  mutate(label = paste0(round_ties_away(pct, 1), "%")) |>
  ggplot(aes(x = pct, y = race_ethn, label = label)) |>
  mh_barplot(fill = "#F04C43", legend = FALSE) +
  scale_x_continuous(expand = c(0, NA)) +
  xlab("\nSample (%)") + ylab("Race/ethnicity\n") #+
# scale_y_discrete(labels = scales::label_wrap(40))

figs$re1
```

```{r}
n <- 648
figcap <- paste0(
  cb$prompts$caption[cb$prompts$number == 1],
  " N = ", n, "."
)

figcap
```

```{r}
mhsamp$race_ethn_all <- list(
  sum = df,
  n = n,
  fig = figs$re1,
  cap = figcap
)
```

## Race/ethnicity (collapsed)

### Summarize

Summarize the survey sample by collapsed race/ethnicity categories.

```{r}
df <- mhsurv |>
  summarize_results(var = "race_ethn_bwo", na_rm = FALSE)
```

```{r}
# Convert to factor sorted by size
lvl <- df |>
  arrange(pct) |>
  pull(race_ethn_bwo)

df <- df |>
  mutate(race_ethn_bwo = factor(race_ethn_bwo, levels = lvl))
```

Pull the ACS population estimates by race for Kansas City.

```{r}
kcpop$race <- acs5_city_2023 |>
  filter(grepl("B01001[A-G]_001", variable)) |>
  mutate(kc_total = kcpop$total)
```

Verify that the sum of estimates by race equals the total population estimate.

```{r}
identical(
  kcpop$race |>
    pull(estimate) |>
    sum(),
  acs5_city_2023 |>
    filter(grepl("B01001_001", variable)) |>
    pull(estimate)
)
```

Summarize the ACS population estimates by the collapsed race/ethnicity categories.

```{r}
kcpop$race_ethn <- data.frame(
  race_ethn = c("Black", "White", "Other"),
  estimate = c(
    kcpop$race$estimate[grepl("Black", kcpop$race$concept)],
    kcpop$race$estimate[grepl("White", kcpop$race$concept)],
    sum(kcpop$race$estimate[!grepl("Black|White", kcpop$race$concept)])
  ),
  kc_total = kcpop$total
)
```

Find the percentage of each collapsed race/ethnicity category in KC sampled.

```{r}
df <- df |>
  left_join(
    kcpop$race_ethn,
    by = c("race_ethn_bwo" = "race_ethn"),
    keep = TRUE # keeps `race_ethn_bwo` as factor
  ) |>
  select(-race_ethn) |>
  mutate(
    label = paste0(round_ties_away(pct), "%"),
    pct_kc = pct(estimate, kc_total, digits = 5),
    d0 = as.numeric(race_ethn_bwo) - d,
    d1 = as.numeric(race_ethn_bwo) + d
  )

df
```

### Plot

Figure: Collapsed race/ethnicity of survey sample. Respondents who chose "White" or "MENA" alone are grouped under "White", "Black" or "African" alone are grouped under "Black", and all others under "Other".

```{r}
figs$rec1 <- df |>
  ggplot(aes(x = pct, y = race_ethn_bwo, label = label)) |>
  mh_barplot(fill = "#F04C43", legend = FALSE) +
  scale_x_continuous(expand = c(0, NA)) +
  xlab("\nSample (%)") + ylab("Race/ethnicity\n")

figs$rec1
```

```{r}
figs$rec2 <- figs$rec1 +
  kc_pop_segment(pct_kc, pct_kc, d0, d1)

figs$rec2
```

```{r}
n <- sum(df$n[!is.na(df$race_ethn_bwo)])
figcap <- paste0(
  "Race/ethnicity (collapsed) of survey sample.",
  " N = ", n, "."
)

figcap
```

```{r}
mhsamp$race_ethn_bwo <- list(
  sum = df,
  n = n,
  fig1 = figs$rec1,
  fig2 = figs$rec2,
  cap = figcap
)
```

## Hispanic origin

### Summarize

Summarize the survey sample by Hispanic origin.

```{r}
df <- mhsurv |>
  summarize_results(var = "hispanic_origin", na_rm = FALSE)
```

Pull the ACS population estimates by Hispanic origin for Kansas City.

```{r}
kcpop$hispanic <- acs5_city_2023 |>
  filter(grepl("B01001I?_001", variable)) |>
  mutate(kc_total = kcpop$total)
```

Summarize the ACS population estimates by Hispanic origin.

```{r}
kcpop$hispanic <- data.frame(
  hispanic_origin_acs = c("Hispanic", "Not Hispanic"),
  estimate = c(
    kcpop$hispanic$estimate[kcpop$hispanic$variable == "B01001I_001"],
    kcpop$hispanic$estimate[kcpop$hispanic$variable == "B01001_001"] -
      kcpop$hispanic$estimate[kcpop$hispanic$variable == "B01001I_001"]
  ),
  kc_total = kcpop$total
)
```

Find the percentage of each Hispanic origin category in KC sampled.

```{r}
df <- df |>
  left_join(
    kcpop$hispanic,
    by = c("hispanic_origin" = "hispanic_origin_acs"),
    keep = TRUE # keeps `hispanic_origin` as factor
  ) |>
  select(-hispanic_origin_acs) |>
  mutate(
    label = paste0(round_ties_away(pct), "%"),
    pct_kc = pct(estimate, kc_total, digits = 5),
    d0 = as.numeric(hispanic_origin) - d,
    d1 = as.numeric(hispanic_origin) + d
  )

df
```

### Plot

Figure: Hispanic origin of survey sample.

```{r}
figs$hisp1 <- df |>
  ggplot(aes(x = pct, y = hispanic_origin, label = label)) |>
  mh_barplot(fill = "#F04C43", legend = FALSE) +
  scale_x_continuous(expand = c(0, NA)) +
  xlab("\nSample (%)") + ylab("Hispanic origin\n")

figs$hisp1
```

```{r}
figs$hisp2 <- figs$hisp1 +
  kc_pop_segment(pct_kc, pct_kc, d0, d1)

figs$hisp2
```

```{r}
print(paste("Answered by", n, "of 648 respondents"))
```

```{r}
n <- sum(df$n[!is.na(df$hispanic_origin)])
figcap <- paste0(
  "Hispanic origin of survey sample.",
  " N = ", n, "."
)

figcap
```

```{r}
mhsamp$hispanic_origin <- list(
  sum = df,
  n = n,
  fig1 = figs$hisp1,
  fig2 = figs$hisp2,
  cap = figcap
)
```

## HOI regions

Summarize survey sample by HOI region.

```{r}
mhsurv |>
  count(hoi_region, .drop = FALSE)
```

Map HOI regions and deidentified respondent residence locations.

```{r fig.width=4, fig.height=6}
ggplot() +
  geom_sf(
    data = sf_city_2023,
    fill = "pink",
    color = "red",
    linewidth = 1
  ) +
  geom_sf(
    data = hoi_regions,
    fill = "lightblue",
    color = "blue",
    linewidth = 1
  ) +
  geom_sf(
    data = mhsurv |>
      st_as_sf(coords = c("block_lon", "block_lat"), crs = 4326),
    color = "black",
    alpha = .5
  )
```

## ZCTA

### Summarize

Find the percentage of each ZCTA population in KC sampled.

```{r}
df <- mhsurv |>
  mutate(zip = factor(zip, levels = sort(as.numeric(sf_zcta_2024$GEOID20)))) |>
  count(zip, .drop = FALSE) |>
  left_join(
    acs5_zcta_2023 |>
      filter(variable == "B01001_001") |>
      select(GEOID, estimate),
    by = c("zip" = "GEOID"),
    keep = TRUE
  ) |>
  select(-GEOID) |>
  mutate(pct_of_zcta = pct(n, estimate, digits = 5))

df
```

### Plot

```{r fig.width=5, fig.height=14}
figs$zcta <- df |>
  mutate(label = paste0(round_ties_away(pct_of_zcta, 2), "%")) |>
  ggplot(aes(x = pct_of_zcta, y = zip, label = label)) |>
  mh_barplot(
    fill = "#F04C43",
    title = "ZCTAs in sample compared to Kansas City population",
    legend = FALSE
  ) +
  scale_x_continuous(expand = c(0, NA)) +
  xlab("\nSample (%)") + ylab("ZCTA\n")

figs$zcta
```

### Choropleth map

```{r}
geo_zcta <- st_transform(sf_zcta_2023, 4326)

kcpop$zcta <- acs5_zcta_2023 |>
  filter(grepl("^B01003", variable))
```

```{r}
pts <- mhsurv |>
  select(id, block_lon, block_lat) |>
  st_as_sf(
    coords = c("block_lon", "block_lat"),
    crs = 4326
  ) |>
  st_join(
    geo_zcta[, c("GEOID20", "geometry")],
    join = st_within
  ) |>
  count(GEOID20, .drop = FALSE) |>
  as.data.frame()
```

```{r}
geo_zcta <- geo_zcta |>
  left_join(
    pts[, c("GEOID20", "n")],
    by = "GEOID20"
  ) |>
  left_join(
    kcpop$zcta[, c("GEOID", "estimate")],
    by = c("GEOID20" = "GEOID")
  ) |>
  rename(n_sampled = n)
```

```{r}
geo_zcta <- geo_zcta |>
  mutate(kcpop = kc_area_pct / 100 * estimate) |>
  mutate(pct_sampled = setmeup::pct(n_sampled, kcpop, 8)) |>
  mutate(n_sampled_per_1k = n_sampled * 1000 / kcpop)
```

```{r fig.width=5, fig.height=6}
figs$geo_zcta <- ggplot(geo_zcta) +
  geom_sf(
    aes(fill = n_sampled_per_1k)
  ) +
  theme_void() +
  scale_fill_viridis_c(
    breaks = c(0, 2, 4, 6, 8),
    name = "Number sampled per\n1,000 residents"
  )

figs$geo_zcta
```

## Map of sample

Map of survey sample with OSM base layer and city boundary.

```{r}
geo_kc <- st_transform(sf_city_2024,  crs = st_crs(3857))
ext <- st_bbox(geo_kc)
exp <- 5000
ext[1] <- ext[1] - exp
ext[2] <- ext[2] - exp
ext[3] <- ext[3] + exp
ext[4] <- ext[4] + exp
```

```{r}
pts <- st_transform(
  mhsurv |>
    select(block_lon, block_lat) |>
    st_as_sf(coords = c("block_lon", "block_lat"), crs = 4326),
  crs = st_crs(3857)
)
```

```{r fig.width=4, fig.height=6}
figs$map_sample <- ggplot(data = geo_kc) +
  basemaps::basemap_gglayer(
    ext = ext,
    map_service = "osm",
    map_type = "streets",
    map_res = 1,
    verbose = FALSE
  ) +
  geom_sf(
    data = geo_kc,
    color = NA,
    fill = "navy",
    alpha = .2
  ) +
  geom_sf(
    data = pts,
    color = "#ccc",
    fill = "navy",
    shape = 21,
    size = 2
  ) +
  coord_sf() +
  scale_fill_identity() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_void()

figs$map_sample
```

```{r}
mhsamp$sample_map <- list(
  fig = figs$map_sample,
  cap = "Map of survey sample. Locations are approximate. N = 648."
)
```

# Save/export

```{r}
saveRDS(mhsamp, "../data/2-final/sample_data_&_plots.rds")
```

```{r}
# ggsave(
#   "../output/mh-report/plot-sex.png",
#   mhsamp$sex$fig1, width = 3.5, height = 2.5, bg = "white"
# )
# ggsave(
#   "../output/mh-report/plot-age.png",
#   mhsamp$age$fig1, width = 3.5, height = 2.5, bg = "white"
# )
# ggsave(
#   "../output/mh-report/plot-pyramid.png",
#   mhsamp$age_sex$fig, width = 7, height = 2.5, bg = "white"
# )
# ggsave(
#   "../output/mh-report/plot-race-ethn-all.png",
#   mhsamp$race_ethn_all$fig, width = 5, height = 2.5, bg = "white"
# )
# ggsave(
#   "../output/mh-report/plot-race-ethn-bwo.png",
#   mhsamp$race_ethn_bwo$fig1, width = 5, height = 2.5, bg = "white"
# )
# ggsave(
#   "../output/mh-report/plot-hispanic.png",
#   mhsamp$hispanic_origin$fig1, width = 5, height = 2.5, bg = "white"
# )
# ggsave(
#   "../output/mh-report/map-respondents.png",
#   mhsamp$sample_map$fig, width = 4.5, height = 7, bg = "white"
# )
```
