---
title: "Additional analyses"
output: html_notebook
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 10,
  fig.height = 5
)
```

```{r message=FALSE}
library(tidyverse)
library(setmeup)
library(kcData)
library(patchwork)
library(gt, exclude = "pct")
```

```{r include=FALSE}
lapply(list.files(pattern = "^fn"), source)
```

```{r}
mhsurv <- readRDS("../data/2-final/mh_survey_results.rds")
slpdur <- readRDS("../data/2-final/sleep_duration_stats.rds")
pop <- readRDS("../data/2-final/kc_population_acs1_2023.rds")
```

# Anxiety & depression

Compare low-frequency and moderate- to high-frequency responses across questions about anxiety and depression.

```{r}
# All responses and numeric values for relevant questions
lapply(mhsurv |> select(q4_1:q7) |> colnames(), \(x) {
  levels(mhsurv[[x]])
})
```

```{r}
# All non-NA responses
df <- mhsurv |>
  filter(if_all(q4_1:q7, ~ !is.na(.x)))

# Low freq
n_low <- df |>
  filter(
    if_all(q4_1:q6, ~ as.numeric(.x) %in% 1:2),
    as.numeric(q7) == 1
  ) |>
  nrow()

# Mod-high freq
n_high <- df |>
  filter(
    if_all(q4_1:q6, ~ as.numeric(.x) %in% 3:5),
    as.numeric(q7) %in% 2:3
  ) |>
  nrow()

# Table
tbl <- data.frame(
  frequency = c("low", "mod-high"),
  n = c(197, 35)
)

tbl$pct <- setmeup::pct(tbl$n, nrow(df), 1)

tbl
```

# Descriptive stats

```{r}
tbl <- list()
```

## Sleep duration

For each age category in the survey sample, find the proportion of the corresponding age category in the KC population, limited to the age range represented in the survey sample.

```{r}
popage <- pop$age |>
  mutate(proportion = estimate / sum(pop$age$estimate))
```

Crude prevalence of inadequate sleep duration prevalence (< 7 hours).

```{r}
mhsurv |>
  drop_na(slp_dur_num) |>
  summarize(pct = pct(sum(slp_dur_num < 7), n())) |>
  pull(pct)
```

Age-adjusted prevalence.

```{r}
df <- mhsurv |>
  rename(age = age1) |>
  drop_na(slp_dur_num, age) |>
  group_by(age) |>
  summarize(pct = pct(sum(slp_dur_num < 7), n(), 5)) |>
  left_join(
    popage |>
      select(age, proportion),
    by = "age"
  ) |>
  mutate(age_adj_rate = pct * proportion)

df
```

```{r}
slpdur$surv <- sum(df$age_adj_rate)

slpdur$surv
```

Create a table with inadequate sleep duration for the survey, as well as the city, state, and US populations (sourced from PLACES). Age-adjusted prevalence is reported from the PLACES data.

```{r}
df <- data.frame(
  population = c(
    "Kansas City (Mental Health Survey)",
    "Kansas City (CDC PLACES)",
    "Missouri (CDC BRFSS)",
    "United States (CDC PLACES)"
  ),
  dur7lt = c(
    slpdur$surv,
    slpdur$kc |>
      filter(grepl("Age-adjust", data_value_type)) |>
      pull(data_value),
    slpdur$mo |>
      pull(age_adjusted_prevalence),
    slpdur$us |>
      filter(grepl("Age-adjust", data_value_type)) |>
      pull(data_value)
  )
)

df <- df |>
  mutate(dur7gt = 100 - dur7lt) |>
  mutate(across(starts_with("slp"), ~ round_ties_away(.x, 0)))

df
```

```{r}
tbl$slp_dur_stats <- gt(df) |>
  fmt_percent(starts_with("dur"), decimals = 0, scale_values = FALSE) |>
  cols_label(
    population = "Population",
    dur7lt = md("Under 7 hours"),
    dur7gt = md("7 hours or more")
  ) |>
  opt_vertical_padding(scale = .25)

tbl$slp_dur_stats
```

## Sleep latency

Descriptive statistics for sleep latency.

```{r}
df <- mhsurv |>
  summarize(
    min = min(slp_lat, na.rm = TRUE),
    max = max(slp_lat, na.rm = TRUE),
    mean = round_ties_away(mean(slp_lat, na.rm = TRUE), 1),
    median = median(slp_lat, na.rm = TRUE),
    mode = mhsurv |>
      count(slp_lat) |>
      arrange(desc(n)) |>
      filter(n == max(n)) |>
      pull(slp_lat) |>
      paste(collapse = ", ")
  ) |>
  ungroup()

df
```

```{r}
# df$msr <- c(
#   "Range of responses",
#   "Average of all responses (mean)",
#   "Center-most response (median)",
#   "Most common response (mode)"
# )

tbl$slp_lat <- df |>
  gt() |>
  cols_label(
    min = "Min.",
    max = "Max.",
    mean = md("Mean<br>(average)"),
    median = md("Median<br>(middle value)"),
    mode = md("Mode<br>(most frequent value)")
  ) |>
  opt_vertical_padding(scale = .25)

tbl$slp_lat
```

# Correlations

## Correlation matrix

<!-- ```{r} -->
<!-- df <- mhsurv |> -->
<!--   select(q4_1:last_col()) |> -->
<!--   select(-starts_with("q11_"), -ends_with("bin")) |> -->
<!--   mutate(across(everything(), as.numeric)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- cormat <- rstatix::cor_mat(df) -->
<!-- ``` -->

## SM use vs SM addiction

```{r}
cor1 <- cor_tests(mhsurv, "sm_use_num", "sm_add_num")
```

```{r}
print(cor1)
```

## SM use vs sleep duration

```{r}
cor2 <- cor_tests(mhsurv, "sm_use_num", "slp_dur_num")
```

```{r}
print(cor2)
```

## SM use vs time to fall asleep

```{r}
cor3 <- cor_tests(mhsurv, "sm_use_num", "slp_lat")
```

```{r}
print(cor3)
```

## SM addiction vs time to fall asleep

```{r}
cor3 <- cor_tests(mhsurv, "sm_add_num", "slp_lat")
```

```{r}
print(cor3)
```

## Reported vs calculated time to fall asleep

```{r}
mhsurv <- mhsurv |>
  mutate(
    slp_lat_calc = (hrs_in_bed - slp_dur_num) * 60,
    slp_lat_diff = slp_lat - slp_lat_calc
  )
```

```{r}
df <- mhsurv |>
  summarize(
    "Non-NA MTS diff" = sum(!is.na(slp_lat_diff)),
    "Unequal MTS" = sum(slp_lat_diff != 0, na.rm = TRUE),
    "MTS diff >= 5" = sum(slp_lat_diff >= 5, na.rm = TRUE),
    "MTS diff >= 15" = sum(slp_lat_diff >= 15, na.rm = TRUE),
    "MTS diff >= 30" = sum(slp_lat_diff >= 30, na.rm = TRUE)
  )

rownames(df) <- "Responses (n)"

t(df)
```

# Save/export

```{r}
gtsave(tbl$slp_lat, "report/img/tbl-slp-lat.png")
gtsave(tbl$slp_dur_stats, "report/img/tbl-slp-dur-stats.png")
```

```{r}
caps <- list(
  slp_lat_stats = "Sleep latency descriptive statistics",
  slp_dur_stats = "Age-adjusted sleep duration in adults"
)
```

```{r}
saveRDS(caps, "../data/2-final/figure_captions_sleep_tables.rds")
```

